Exploring environmental impacts of colonial contacts on Indigenous communities in northeastern Taiwan

### 2024-07-17 analysis/paper/paper.qmd
work on the section of background

# Introduction
<!-- introduction about European expansion -->
European expansion since the 15th century marked the beginning of an era characterized by global trade, exploration, and colonization, which often led to profound impacts on indigenous societies in many parts of the world. Due to specific spatial and temporal contexts, the forms of European colonization and power relations varied across regions [@Jorden2009]. Understanding the dynamics of these colonial processes has become a significant issue in archaeology, as it provides critical insights into the complex outcomes of European colonization [@Gosden2004]. In Asia, European colonization was primarily driven by the pursuit of Asian luxury goods and local natural resources for commercial profit. The new exchange systems and governance policies imposed by European colonizers often had profound effects on local economic systems, social structures, and settlement patterns [@Christie1996; @Junker2004; @Lape2002]. However, Indigenous communities did not just passively accept European colonization. Recent discussions on colonial interactions also highlight the diverse strategies employed by Indigenous communities in response to European colonization, such as resistance, negotiation, incorporation, persistence, or resilience [@Torrence2003; @Silliman2005; @Mitchell2010].

<!-- NE Taiwan as a case study -->
While many studies focus on the center of European colonies, places on the periphery of colonial center were also more and less impacted by European powers or associated trading systems [@Acabado2017; @Trabert2017]. Exploring these areas is important to gain a broader understanding of the varied consequences of colonial expansion. Northeastern Taiwan, specifically Yilan, serves as an example of a peripheral case in the 17th century when Europeans established their colonial base in northern Taiwan. Due to its geologically challenging and relatively inaccessible location, Yilan experienced limited direct European control. However, local settlements were frequently involved in the regional exchange network stimulated by the colonial presence in the north. In the mid-17th century, the VOC (Dutch East India Company) formalized the deerskin trade in Taiwan for export, involving many Indigenous communities and Han Chinese. Despite some discussions on the policies and operations of the deerskin trade, direct archaeological investigations into the reactions of local Indigenous people to these commercial activities are rare. Prior to the deerskin trade, deer were already a substantial part of local subsistence. Whether the interaction between Indigenous people and their environment changed due to European trade demands requires further archaeological investigation.

<!-- research aim and questions -->
In this paper, I will explore the interactions between people and their environment in a colonial context by examining archaeological evidence from the Kiwulan site (1350-1950 CE), a large Indigenous settlement in northeastern Taiwan [@citation]. Kiwulan experienced periods of European presence in northern Taiwan in the 17th century, followed by a large influx of Han Chinese migration in the 19th century. The later period of Chinese migration will also be included to provide a long-term perspective. Archaeological evidence related to subsistence practices is particularly focused in this research, as the choices made in these practices can reveal the strategies Indigenous people employed to cope with European colonial activities and the associated changes in their environment. My research questions are: What were the resource use patterns of Indigenous people before and after the European presence? Can we identify shifts in these patterns in response to colonial economic demands? To address these questions, I will examine the faunal and plant remains from Kiwulan, alongside a preliminary analysis of ceramic residues. By adopting an environmental and ecological perspective, this study aims to shed light on Indigenous persistence and resilience strategies in the face of foreign interruptions.

# Historical background
As an island close to the Asian continent, Northeast Asia, and Island Southeast Asia, Taiwan has been a place with frequent cross-cultural interactions since the Neolithic period and involved in a complex global trading network in the 17th century due to the European expansion. Based on the commercial purposes, the Europeans, including the Spanish and the Dutch, built fortresses and trading posts at different parts of Taiwan.

@@

# colonial studies in Archaeology

@@

# Background
The Yilan Plain is the major coastal plain in northeastern Taiwan. Prehistory in the Yilan Plain begins from the Middle Neolithic period around 4000 B.P. and continues throughout the Metal Age period (2000-400 B.P.) (Liu 2011). The final Late Iron Age in Taiwan can be viewed as a protohistoric period due to the interaction of record-keeping Chinese and Europeans with local cultures. Most Late Metal Age sites in the Yilan Plain can be related to the local Indigenous people, the Kavalan, according to oral history and similarities between historical Kavalan material culture and archaeological artifacts (Chen 2004). 
European colonization of Taiwan began in the 1620s because of the competition between the Spanish and the Dutch for political and economic control of the Pacific (Andrade, 2007). Yilan is about 70 kilometers away from the first European northern base with mountains in between. Indigenous people in Yilan were not in direct contact with colonial powers until 1632, when they were attacked by the Spanish. Later in 1647, the Dutch attacked the Indigenous people and forced them to accept colonial rules and economic demands by paying annual tribute (Andrade, 2007). The Dutch also built a trading outpost in Trobiawan. According to the Dutch census, Indigenous people in northeastern Taiwan occupied over 40 settlements with approximately 10,000 people in the 17th century (Nakamura 1938), and those settlements were also described by the Chinese in later periods, demonstrating a continuous Indigenous habitation until the 19th century (Figure 1).
Due to the fear of European attacks, most local communities in Yilan maintained good relationships with the Europeans by paying tributes and Indigenous leaders could obtain foreign materials as gifts according to the degree of allegiance (Kang 2012). In addition to historical records, archaeological evidence also shows complex inter- and intra- exchange networks based on the findings of foreign goods, such as stonewares commonly found in European outposts in Southeast Asia, Chinese porcelains, and glass and stone beads (Chen 2005; Chen 2007; Liu 2011; National Museum of Taiwan History 2005, Wang and Liu 2007). This suggests frequent trade activities since the protohistoric period in the 14th century and later intensified through the global trade network stimulated by the Europeans and associated colonial activities. Moreover, foreign materials such as ceramics, beads, and textile were viewed as prestige goods in local communities before European arrival (Li and Wu, 2006).

# Methods
## Geoarchaeological and geochemical analysis for sediments
To explore changes in environments, I will examine sediments from archaeological sites in northeastern Taiwan, Yilan. I will mainly focus on archaeological sites recorded as Indigenous settlements that can be traced back to the protohistoric and historic period when they experienced the European and Chinese presence, such as Kiwulan I, Kiwulan II, and Yilan Agricultural Vocational High School site. Kiwulan and Yilan Agricultural Vocational High School are two major Late Metal Age sites with extensive excavation in recent years. Kiwulan (600-100 B.P.) is located at the northern part of Yilan plain (Chen, 2007), while Yilan Agricultural Vocational High School site (400-200 B.P.) is situated on the central part of Yilan (Li, 2014). Archaeological evidence from both sites includes a large amount of pottery sherds, imported stoneware and porcelains, wooden artifacts, stone and metal tools, pipes, thousands of glass beads, agate beads and metal ornaments, post holes with in-situ wooden posts, and burials. Based on their locations and the historical records, those two sites can be traced back to the 17th century when they were named Banouran and Beili according to the Dutch census (Nakamura 1938). Those archaeological materials are mainly stored in the Cultural Affairs Bureau in Yilan County, and it is feasible to analyze those materials and collect data based on previous collaboration during my PhD. 
For archaeological sediments, I will use geoarchaeological methods to measure the physical and chemical properties of sediments, including particle size, magnetic susceptibility, pH values, electrical conductivity, and organic matter and carbonates proportions to understand soil conditions and Indigenous landuse. I will also use organic chemical analysis to detect lipids in sediments to reconstruct past environments, fire and land use patterns associated with the colonial presence and long-term cultural interaction. Geoarchaeological and geochemical methods are useful and insightful methods especially when faunal and botanical evidence were poorly preserved, such as in a condition of acidic environments.

>>>>

# Introduction
<!-- introduction about European expansion -->
European expansion since the 15th century marked the beginning of an era characterized by global trade, exploration, and colonization, which often led to profound impacts on indigenous societies in many parts of the world. According to specific spatial and temporal contexts, the forms of European colonization and power relations varied across regions [@Jorden2009]. Understanding the dynamics of these colonial processes has become a significant issue in archaeology, as it provides critical insights into the complex outcomes of European colonization [@Gosden2004]. In Asia, European colonization was primarily driven by the procurement of Asian luxury goods, raw materials, or natural resources for commercial profit. The new exchange systems and governance policies imposed by European colonizers often had profound effects on local economic systems, social structures, and settlement patterns [@Christie1996; @Junker2004; @Lape2002]. However, Indigenous communities did not just passively accept European colonization. Recent discussions on colonial interactions also highlight the diverse strategies employed by Indigenous communities in response to European presence, such as resistance, negotiation, persistence, resilience, or entanglement [@Torrence2003; @Silliman2005; @Mitchell2010; @Jorden2009].

<!-- NE Taiwan as a case study -->
While many studies focus on prominent colonies, places on the periphery of colonial center were also more and less impacted by European powers or associated trading systems [@Acabado2017; @Trabert2017]. Exploring these areas is important to gain a broader understanding of the varied consequences of colonial expansion. Northeastern Taiwan, specifically Yilan, serves as an example of a peripheral case in the 17th century when Europeans established their colonial base in northern Taiwan. Due to its geologically challenging and relatively inaccessible location, Yilan experienced limited direct European control. However, local settlements were frequently involved in the regional exchange network stimulated by the colonial presence in the north. In the mid-17th century, the VOC (Dutch East India Company) formalized the deerskin trade in Taiwan for export, involving many Indigenous communities and Han Chinese. Despite some discussions on the policies and operations of the deerskin trade, direct archaeological investigations into the experiences of local Indigenous people to these commercial activities are rare. The European impacts on local ecological conditions have been recorded in many cases in colonial settings [@Jorden2009]. Prior to the deerskin trade, deer has been a substantial part of local subsistence. Whether the interaction between Indigenous people and their environment changed due to European trade demands requires further archaeological investigation.

<!-- research aim and questions -->
In this paper, I will explore the interactions between people and their environment in a colonial context by examining resource use of indigenous people through a case study at the Kiwulan site (1350-1950 CE), a largest settlement in northeastern Taiwan [@citation]. Kiwulan experienced periods of European presence in the 17th century, followed by a large influx of Han Chinese migration in the 19th century. I will primarily focus on subsistence practices, as the choices made in these practices can reveal the strategies Indigenous people employed to cope with European colonial activities and the associated changes in their environment. My research questions are: What were the subsistence practices of Indigenous people before and after the European presence? Can we identify shifts in these strategies in response to colonial economic demands? To address these questions, I will examine the faunal and plant remains from Kiwulan, alongside a preliminary analysis of ceramic residues for investigating culinary practices. Through an environmental perspective, this study aims to shed light on Indigenous persistence and resilience strategies in the face of foreign interruptions.

# Colonial encounter in Northeastern Taiwan
<!-- the structure of power relations, colonial control in particular spatial and temporal contexts-->
In 1626, the Spanish colonized northern Taiwan and built fortress in Tamsui and Keelung to avoid the interference in the Asian trade by the VOC (Vereenigde Oost-Indische Compagnie; Dutch East India Company), which established a colony in southern Taiwan at the same time. In 1642, the Dutch defeated the Spanish and took over their fortresses, officially expanding their colonial power to northern Taiwan until the end of their colonization in 1662 [@Andrade2007]. The form of European colonization in northern Taiwan relied to a certain degree on local Indigenous people for the development of the colony and the collection resources they needed among different Indigenous groups. For example, the Basay people, active in Keelung, were recorded to have assisted in the colonial activities of the VOC administrators, such as coordinating between indigenous villages [@citation]. The European domination was mostly established through the conquest over indigenous villages, and the VOC further developed tribute system to request deerskins or products from indigenous people []. However, for most indigenous villages distant from the European colonial base, European colonial control was relatively less, indirect, and even unsuccessful.

<!-- domination and the processes by which it was established and resisted, how and why colonial powers were able to exert control (however limited) over Native populations -->
Kavalan, or Yilan in the present day, in northeastern Taiwan is an example of slight European presence for most of the time. Kavalan is about 70 kilometers away from the European northern base with mountains in between. Local indigenous groups were Kavalan people, who first experienced European powers in 1632, when they were attacked by the Spanish. Later in 1647, a group of Dutch soldiers attacked many Kavalan villages and asked those Kavalan villages to pay tributes [@citation]. The Dutch also built a trading outpost in Trobiawan located near the coastal area at Yilan, but there might be not much officials at any given time. According to the Dutch census, there were more than 40 Kavalan villages with approximately 10,000 people in the 17th century in this region [@Nakamura1938], which were also mentioned by the Chinese and Japanese in later periods, demonstrating a continuous Indigenous habitation until the 19th century. It appears that the Kavalan area was in the situation where trading activities were frequent and active stimulated by the European presence in northern Taiwan, with occasional pressures from colonial powers. In general, the European control over the Kavalan must have been largely indirect with limited colonialism [@].

Before the arrival of the Europeans in northern Taiwan,

The Yilan Plain is the major coastal plain in northeastern Taiwan. Prehistory in the Yilan Plain begins from the Middle Neolithic period around 4000 B.P. and continues throughout the Metal Age period (2000-400 B.P.) (Liu 2011). The final Late Iron Age in Taiwan can be viewed as a protohistoric period due to the interaction of record-keeping Chinese and Europeans with local cultures. Most Late Metal Age sites in the Yilan Plain can be related to the local Indigenous people, the Kavalan, according to oral history and similarities between historical Kavalan material culture and archaeological artifacts (Chen 2004).

 Yilan is about 70 kilometers away from the first European northern base with mountains in between. Indigenous people in Yilan were not in direct contact with colonial powers until 1632, when they were attacked by the Spanish. Later in 1647, the Dutch attacked the Indigenous people and forced them to accept colonial rules and economic demands by paying annual tribute (Andrade, 2007). The Dutch also built a trading outpost in Trobiawan. According to the Dutch census, Indigenous people in northeastern Taiwan occupied over 40 settlements with approximately 10,000 people in the 17th century (Nakamura 1938), and those settlements were also described by the Chinese in later periods, demonstrating a continuous Indigenous habitation until the 19th century (Figure 1).
Due to the fear of European attacks, most local communities in Yilan maintained good relationships with the Europeans by paying tributes and Indigenous leaders could obtain foreign materials as gifts according to the degree of allegiance (Kang 2012). In addition to historical records, archaeological evidence also shows complex inter- and intra- exchange networks based on the findings of foreign goods, such as stonewares commonly found in European outposts in Southeast Asia, Chinese porcelains, and glass and stone beads (Chen 2005; Chen 2007; Liu 2011; National Museum of Taiwan History 2005, Wang and Liu 2007). This suggests frequent trade activities since the protohistoric period in the 14th century and later intensified through the global trade network stimulated by the Europeans and associated colonial activities. Moreover, foreign materials such as ceramics, beads, and textile were viewed as prestige goods in local communities before European arrival (Li and Wu, 2006).

@@

@@

# Methods
The later period of Chinese migration will also be included to provide a long-term perspective

## Geoarchaeological and geochemical analysis for sediments
To explore changes in environments, I will examine sediments from archaeological sites in northeastern Taiwan, Yilan. I will mainly focus on archaeological sites recorded as Indigenous settlements that can be traced back to the protohistoric and historic period when they experienced the European and Chinese presence, such as Kiwulan I, Kiwulan II, and Yilan Agricultural Vocational High School site. Kiwulan and Yilan Agricultural Vocational High School are two major Late Metal Age sites with extensive excavation in recent years. Kiwulan (600-100 B.P.) is located at the northern part of Yilan plain (Chen, 2007), while Yilan Agricultural Vocational High School site (400-200 B.P.) is situated on the central part of Yilan (Li, 2014). Archaeological evidence from both sites includes a large amount of pottery sherds, imported stoneware and porcelains, wooden artifacts, stone and metal tools, pipes, thousands of glass beads, agate beads and metal ornaments, post holes with in-situ wooden posts, and burials. Based on their locations and the historical records, those two sites can be traced back to the 17th century when they were named Banouran and Beili according to the Dutch census (Nakamura 1938). Those archaeological materials are mainly stored in the Cultural Affairs Bureau in Yilan County, and it is feasible to analyze those materials and collect data based on previous collaboration during my PhD. 
For archaeological sediments, I will use geoarchaeological methods to measure the physical and chemical properties of sediments, including particle size, magnetic susceptibility, pH values, electrical conductivity, and organic matter and carbonates proportions to understand soil conditions and Indigenous landuse. I will also use organic chemical analysis to detect lipids in sediments to reconstruct past environments, fire and land use patterns associated with the colonial presence and long-term cultural interaction. Geoarchaeological and geochemical methods are useful and insightful methods especially when faunal and botanical evidence were poorly preserved, such as in a condition of acidic environments.
